---
title: "Probabilistic Regime-Switching with Stan"
output: html_document
date: "2024-10-01"
---

```{r setup, include=FALSE}
# Setup for R Markdown
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This R Markdown document demonstrates a probabilistic regime-switching model using Stan. We generate synthetic data, fit a regime-switching model using Stan, and visualize the results.

## 1. Load Required Libraries

```{r rs_bhs_soft_1}
# Load necessary libraries
library(ggplot2)
library(gridExtra)
library(rstan)  # For using Stan
```

## 2. Generate Synthetic Data
We will now generate synthetic time-series data with probabilistic regime-switching behavior.

```{r rs_bhs_soft_2}
# Set seed for reproducibility
set.seed(123)

# Number of time points
n <- 100 

# Initialize variables
X <- numeric(n)  # Hidden state
Y <- numeric(n)  # Observed data
state <- numeric(n)  # Regime states (1 or 2)

# Initial state
state[1] <- 1
X[1] <- rnorm(1, 0, 1)  # Initial hidden state

# Generate data based on regime-switching
for (t in 2:n) {
  if (X[t-1] > 0) {
    state[t] <- 1  # Regime 1
    X[t] <- 0.5 * X[t-1] + rnorm(1, 0, 1)
  } else {
    state[t] <- 2  # Regime 2
    X[t] <- -0.8 * X[t-1] + rnorm(1, 0, 1)
  }
  # Observation model with noise
  Y[t] <- 1.0 * X[t] + rnorm(1, 0, 0.5)
}
```

## 3. Define and Compile Stan Model
We will define a Stan model to estimate the hidden states and threshold for regime-switching.

```{r rs_bhs_soft_3}
# Define Stan model code
stan_code <- "
data {
  int<lower=1> N;       // Number of observations
  vector[N] Y;          // Observed data
}
parameters {
  vector[N] X;          // Hidden states
  real<lower=0> sigma;  // Measurement noise
  real alpha;           // Regime 1 coefficient
  real beta;            // Regime 2 coefficient
  real<lower=0> tau;    // Transition noise
  real threshold;       // Threshold for switching regimes
}
model {
  // Priors
  alpha ~ normal(0, 1);
  beta ~ normal(0, 1);
  sigma ~ normal(0, 1);
  tau ~ normal(0, 1);
  threshold ~ normal(0, 1);

  // Initial hidden state
  X[1] ~ normal(0, 1);

  // Regime-switching dynamics
  for (t in 2:N) {
    real p;
    p = inv_logit(X[t-1] - threshold);  // Probabilistic regime
    X[t] ~ normal((p * alpha + (1 - p) * beta) * X[t-1], tau);
  }

  // Observation model
  Y ~ normal(X, sigma);
}
"

# Compile the Stan model
stan_model <- stan_model(model_code = stan_code)
```

## 4. Fit the Model with Stan
```{r rs_bhs_soft_4}
# Prepare data for Stan
stan_data <- list(N = n, Y = Y)

# Fit the model using Stan
fit <- sampling(stan_model, data = stan_data, iter = 2000, chains = 4)

# Extract hidden states and estimated threshold
hidden_states <- extract(fit)$X
threshold_estimated <- extract(fit)$threshold

# Calculate probabilities of being in regime 1
p_estimated <- 1 / (1 + exp(-(apply(hidden_states, 2, mean) - mean(threshold_estimated))))
```

## 5. Visualize Results
We will now visualize the actual and predicted hidden states along with regime probabilities.
```{r rs_bhs_soft_5}
# Prepare data for visualization
df_hidden <- data.frame(
  time = 1:n, 
  X_estimated = apply(hidden_states, 2, mean),
  p_estimated = p_estimated
)

df <- data.frame(
  time = 1:n, 
  X = X, 
  Y = Y, 
  state = factor(state)
)

# Plot actual hidden states with regime coloring
plot_X_actual <- ggplot(df, aes(x = time, y = X, color = state, group = 1)) +
  geom_line(size = 1) +
  labs(title = "Time vs Actual X (Regime Switching)", x = "Time", y = "X") +
  scale_color_manual(values = c("red", "blue"), name = "Regime")

# Plot predicted hidden states with probabilistic regime switching
plot_X_predicted <- ggplot(df_hidden, aes(x = time, y = X_estimated, color = p_estimated, group = 1)) +
  geom_line(size = 1) +
  labs(title = "Time vs Predicted X (Probabilistic Regime Switching)", x = "Time", y = "X_estimated") +
  scale_color_gradient(low = "red", high = "blue", name = "P(Regime 1)")

# Display plots side by side
grid.arrange(plot_X_actual, plot_X_predicted, ncol = 1)
```